---
title: "Denfensive R"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Conflict

check for new versions of installed packages
```{r}
old <- old.packages()
update.packages(ask = F)
```

### Packages and namespace

packages use namespaces as spaces for names
```{r}
library(dplyr) # give direct access to exported functions in dplyr
# over 200 
getNamespaceExports("dplyr")
```

Gives the order R will look for the function you called
```{r}
search()
library("conflicted") # make function choice explicitly
filter # search this function

# Prefer the dplyr version of the lag function
conflict_prefer("lag", "dplyr")
```

## Early warning

Type TRUE and FALSE instead of T or F, because TRUE is protected but T can be overwritten
use isTRUE() to test if a conditin is true
```{r}
isTRUE(T)
isTRUE(2)
```

### message()
get message(): helpful information about the progress but can be noisy
helpful if running long process
```{r}
noisy <- function(a, b) {
  message("I'm doing stuff")
  a + b
}
noisy(3, 5) # noisy way
suppressMessages(noisy(3,5)) # mute message

suppressPackageStartupMessages(library("tidyverse")) # mute messages when loading package
```

write simulation function
```{r}
# Wrap the code below in a function get_distribution()
# Modify the function to make it less noisy
get_distribution <- function(N, verbose = TRUE) {
  results <- numeric(N)
  for(i in 1:N) {
    results[i] <- simulate()
    # Check if verbose is TRUE
    if(isTRUE(verbose)) {
      # Show a progress report
      message("Simulation ", i, " completed")
    }
  }
  return(results)
}
```


### warning

R will continue working although warning() telling you something is wrong
```{r}
# Create new variable 
x <- c(1, 1, 1)
y <- 1:3

# Correlation coefficient of x & y
cor(x, y)
suppressWarnings(cor(x, y))
```

### stop()

stop the program and print error message 
```{r}
stop("print the error message here")
```

Example
```{r}
mean_age = function(ages) {
  if(any(ages < 0)) {
    stop("You have negative ages!")
  }
  # Stop the execution if any of the ages are over 150
  if(any(ages > 150)) {
    stop("that's impossible!")
  }
  m = mean(ages)
  return(m)
}
```


```{r}
results <- try("Scotland"+"world cup", silent = TRUE)
class(results)

if(class(results) == "try-error") {
  # do something
}
```

### dot (full stop)

try to avoid naming object with "." variable name, because 
```{r}
summary(lm_result)  # S3 object
summary.lm(lm_result) # this is what really happend in the backend for S3 object
```

