---
title: "Chapter2_names&values"
author: "Eleanor Zhang"
date: "2/21/2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(easypackages) # to load multiple packages
libraries("tidyverse", "pryr", "lobstr")
```

## Names vs. values

Tips:  
1) try to avoid extra copies

```{r}
df <- data.frame(runif(3), runif(3))
names(df) <- c(1, 2)
df
df$`3` <- df$`1` + df$`2`
```

Memory used
```{r}
x <- runif(1e6)
y <- list(x, x, x)

object_size(x)
object_size(y) # used up the same memory as a single x
```

When a is copied
```{r}
a <- c(1, 5, 3, 2)
b <- a
b[[1]] <- 10 # a is copied here
```

## 2.2 Binding

x and y share the same location/address; this address changes when you restart R
```{r}
x <- c(1, 2, 3)
y <- x

obj_addr(x)
#> [1] "0x1cde3d8"
obj_addr(y)
#> [1] "0x1cde3d8"
```

## copy and modify behavior

R objects are unchangeable

```{r}
x <- c(1, 2, 3)
cat(tracemem(x), "\n") # start tracing address change

y <- x
y[[3]] <- 4L

y[[3]] <- 5L

untracemem(x) # stop tracing
```

pointers in function calls
```{r}
f <- function(a) {
  a
}

x <- c(1, 2, 3)
cat(tracemem(x), "\n")
#> <0x8f5c7f8>

z <- f(x)
# there's no copy here!

untracemem(x)
```

list: each element in the list point to a name; so it stores both values and their references as well. It also use copy and modify, but it is a shallow copy, meaning only the list object and its bindings (names) are copied, not the values (compared to vectors, where values/contents are copied as well)

Dataframe: list of vectors (can be very messy..)

Character vectors: put elements in a global string pool

compare these two ??? why different from result output
```{r}
x <- c(1L, 2L, 3L)
tracemem(x)

x[[3]] <- 4
untracemem(x) 

#------
x <- c(1, 2, 3)
cat(tracemem(x), "\n")
y <- x
y[[3]] <- 4L
y[[3]] <- 5L

untracemem(x)
```

## Object size

```{r}
x <- runif(1e6)
obj_size(x)
#> 8,000,048 B

y <- list(x, x, x)
obj_size(y)
object_size(y)
#> 8,000,128 B
```

if there is shared values between objects, then the object size of combining them will not be equal to the sum of individual

alternative representation
```{r}
obj_size(1:3)
#> 680 B
obj_size(1:1e3)
#> 680 B
obj_size(1:1e6)
#> 680 B
obj_size(1:1e9)
#> 680 B
```

## 2.5 modify in place

an object has single name bound to it; name = object
```{r}
v <- c(1, 2, 3)
cat(tracemem(x), "\n")

v[[3]] <- 4 # no copy
untracemem(x)
```

Example: for loop

```{r}
x <- data.frame(matrix(runif(5 * 1e4), ncol = 5))
cat(tracemem(x), "\n")
medians <- vapply(x, median, numeric(1))

for (i in seq_along(medians)) {
  x[[i]] <- x[[i]] - medians[[i]]
}

untracemem(x)
```

we can modify this by using list
```{r}
y <- as.list(x)
cat(tracemem(y), "\n")
#> <0x7f80c5c3de20>
  
for (i in 1:5) {
  y[[i]] <- y[[i]] - medians[[i]]
}
#> tracemem[0x7f80c5c3de20 -> 0x7f80c48de210]: 
```

Environments: always modified in place
```{r}

```

