---
title: "Data Manipulation with dplyr"
output: html_document
---

dplyr package: manipulate data very fast

```{r}
# Load the dplyr package
library(dplyr)
# Load the hflights package
library(hflights)

head(hflights)
summary(hflights)
glimpse(hflights)
```

### Convert data frame to tibble:  
The `tbl` format changes how R displays your data, but it does not change the data's underlying data structure. A tbl inherits the original class of its input, in this case, a data.frame.

```{r}
hflights <- tbl_df(hflights)
hflights
```


#### change labels of data
toy example: 
```{r}
two <- c("AA", "AS")
lut <- c("AA" = "American", 
         "AS" = "Alaska", 
         "B6" = "JetBlue")
two <- lut[two]
two
```

change labels of flight carriers 
```{r}
#look up table
lut <- c("AA" = "American", "AS" = "Alaska", "B6" = "JetBlue", "CO" = "Continental", 
         "DL" = "Delta", "OO" = "SkyWest", "UA" = "United", "US" = "US_Airways", 
         "WN" = "Southwest", "EV" = "Atlantic_Southeast", "F9" = "Frontier", 
         "FL" = "AirTran", "MQ" = "American_Eagle", "XE" = "ExpressJet", "YV" = "Mesa")

# Add the Carrier column to hflights
hflights$Carrier <- lut[hflights$UniqueCarrier]

# Glimpse at hflights
glimpse(hflights)
```

#### five verbs of dplyr and their meaning:

* `select()`: return a subsect of __columns__
* `filter()`: return a subset of __rows__
* `arrange()`: reorder the rows accroding to the variables, argument order matters!
* `mutate()`: add columns to existing data
* `summarise()`: reduce each group into a single row and calculate aggregate measures.

```{r}
hflights %>% filter(UniqueCarrier == "AA") %>% summarise(mean(ArrDelay))
```

practice with `select` to select columns
```{r}
# Print out a tbl with the four columns of hflights related to delay
select(hflights, ActualElapsedTime, AirTime, ArrDelay, DepDelay)

# Print out the columns Origin up to Cancelled of hflights
select(hflights, Origin:Cancelled)

# Answer to last question: be concise!
select(hflights, Year:DayOfWeek, ArrDelay:Diverted) # select multiple cols
```

helper function that come with `select`:

* `starts_with("X")`: every name starts with "X"
* `ends_with("X")`: every name ends with "X"
* `contains("X")`: every name contains "X"
* `matches("X")`: every name that matches "X", where "X" can be a regular expression
* `num_range("x", 1:5)`: the variables named x01, x02, x03, x04 and x05
* `one_of(x)`: every name that appears in x, which should be a character vector.

Practice:
```{r}
# Print out a tbl containing just ArrDelay and DepDelay
select(hflights, contains("Delay"))

# Print out a tbl as described in the second instruction, using both helper functions and variable names
select(hflights, UniqueCarrier, contains("Num"), contains("Cancel"))

# Print out a tbl as described in the third instruction, using only helper functions.
select(hflights, contains("Time"), contains("Delay"))
```

More practice:
```{r}
# Finish select call so that ex1d matches ex1r
ex1d <- select(hflights, TaxiIn, TaxiOut, Distance)
ex1d

# Finish select call so that ex2d matches ex2r
ex2d <- select(hflights,Year:DayOfWeek, -DayofMonth, DepTime, ArrTime)
ex2d
# Finish select call so that ex3d matches ex3r
ex3d <- select(hflights, starts_with("Ta"))
ex3d
```

#### mutating
```{r}
# Add the new variable ActualGroundTime to a copy of hflights and save the result as g1.
g1 <- mutate(hflights, ActualGroundTime = ActualElapsedTime - AirTime)
glimpse(g1)

# Add the new variable GroundTime to g1. Save the result as g2.
g2 <- mutate(g1, GroundTime = TaxiIn + TaxiOut)
glimpse(g2)

# Add the new variable AverageSpeed to g2. Save the result as g3.
g3 <- mutate(g2, AverageSpeed = Distance / AirTime * 60)
glimpse(g3)
```


```{r}
# Add a second variable loss_ratio to the dataset: m1
m1 <- mutate(hflights, loss = ArrDelay - DepDelay, loss_ratio=loss/DepDelay)
glimpse(m1)

# Add the three variables as described in the third instruction: m2
m2 <- mutate(hflights, TotalTaxi = TaxiIn + TaxiOut,ActualGroundTime = ActualElapsedTime - AirTime, Diff = TotalTaxi - ActualGroundTime)

glimpse(m2)
```

#### logical operators

* x < y, TRUE if x is less than y
* x <= y, TRUE if x is less than or equal to y
* x == y, TRUE if x equals y
* x != y, TRUE if x does not equal y
* x >= y, TRUE if x is greater than or equal to y
* x > y, TRUE if x is greater than y
* `x %in% c(a, b, c)`, TRUE if x is in the vector c(a, b, c)

```{r}
# All flights that traveled 3000 miles or more
hflights %>% filter(Distance>3000)

# All flights flown by one of JetBlue, Southwest, or Delta
hflights %>% filter(Carrier %in% c("JetBlue","Southwest","Delta"))

# All flights where taxiing took longer than flying
hflights %>% filter(TaxiIn + TaxiOut > AirTime)
```


Practice on Boolean
```{r}
# All flights that departed before 5am or arrived after 10pm
filter(hflights, DepTime < 500 |ArrTime > 2200)

# All flights that departed late but arrived ahead of schedule
filter(hflights, DepDelay > 0, ArrDelay < 0)

# All flights that were cancelled after being delayed
filter(hflights, DepDelay > 0, Cancelled == 1)
```

Blend practice
```{r}
# Select the flights that had JFK as their destination: c1
c1 <- filter(hflights, Dest == "JFK")
c1

# Combine the Year, Month and DayofMonth variables to create a Date column: c2
c2 <- mutate(c1, Date = paste(Year, Month, DayofMonth, sep = "-"))
glimpse(c2)

# Print out a selection of columns of c2
select(c2, Date, DepTime, ArrTime, TailNum)
```

#### arrange
by default, `arrange` will sort data from smallest to largest.

```{r}
# Definition of dtc
dtc <- filter(hflights, Cancelled == 1, !is.na(DepDelay))
dtc

# Arrange dtc by departure delays
arrange(dtc, DepDelay)

# Arrange dtc so that cancellation reasons are grouped
arrange(dtc, CancellationCode)

# Arrange dtc according to carrier and departure delays
arrange(dtc,UniqueCarrier, DepDelay)
```

reverse the order by `desc()`
```{r}
# Arrange according to carrier and decreasing departure delays
arrange(hflights, UniqueCarrier, desc(DepDelay))

# Arrange flights by total delay (normal order).
arrange(hflights, DepDelay + ArrDelay)
```

#### Summarise
`summarise()` does not return an altered copy of the dataset it is summarizing; instead, it builds a new dataset that contains only the summarising statistics: it turns a row into a single value.

```{r}
# Print out a summary with variables min_dist and max_dist
summarise(hflights, max_dist = max(Distance), min_dist = min(Distance))

# Print out a summary with variable max_div: the max distance of diverted flights
filter(hflights, Diverted == 1) %>% summarise(max_div = max(Distance))
```

Aggregate functions

* min(x) - minimum value of vector x.
* max(x) - maximum value of vector x.
* mean(x) - mean value of vector x.
* median(x) - median value of vector x.
* quantile(x, p) - pth quantile of vector x.
* sd(x) - standard deviation of vector x.
* var(x) - variance of vector x.
* IQR(x) - Inter Quartile Range (IQR) of vector x.
* diff(range(x)) - total range of vector x.

```{r}
# Remove rows that have NA ArrDelay: temp1
temp1 <- filter(hflights, !is.na(ArrDelay))
temp1

# Generate summary about ArrDelay column of temp1
temp1 %>% summarise(earliest = min(ArrDelay), average = mean(ArrDelay), latest = max(ArrDelay), sd = sd(ArrDelay))

# Keep rows that have no NA TaxiIn and no NA TaxiOut: temp2
temp2 <- filter(hflights, !is.na(TaxiIn), !is.na(TaxiOut))
temp2

# Print the maximum taxiing difference of temp2 with summarise()
summarise(temp2, max_taxi_diff = max(abs(TaxiIn - TaxiOut)))
```

dplyr aggregate function:

* first(x) - The first element of vector x.
* last(x) - The last element of vector x.
* nth(x, n) - The nth element of vector x.
* n() - The number of rows in the data.frame or group of observations that summarise() describes.
* n_distinct(x) - The number of unique values in vector x

```{r}
# Generate summarizing statistics for hflights
summarise(hflights, n_obs = n(),n_carrier = n_distinct(UniqueCarrier),n_dest = n_distinct(Dest))

# All American Airline flights
aa <- filter(hflights, UniqueCarrier == "AA")

# Generate summarizing statistics for aa 
summarise(aa, n_flights = n(), n_canc = sum(Cancelled == 1), avg_delay = mean(ArrDelay, na.rm = TRUE))
```

#### pipe

```{r}
# Write the 'piped' version of the English sentences.
hflights %>% mutate(diff = TaxiOut - TaxiIn) %>% filter(!is.na(diff)) %>% summarize(avg = mean(diff))
```

chain together:
```{r}
# Chain together mutate(), filter() and summarise()
hflights %>% 
  mutate(RealTime = ActualElapsedTime + 100, mph = Distance/RealTime*60) %>% 
  filter(!is.na(mph), mph < 70) %>%       
  summarise(n_less = n(), n_dest = n_distinct(Dest), min_dist = min(Distance), max_dist= max(Distance))
```


```{r}
# Make an ordered per-carrier summary of hflights
hflights %>%
   group_by(UniqueCarrier) %>%
   summarise(p_canc = mean(Cancelled == 1) * 100, 
             avg_delay = mean(ArrDelay, na.rm = TRUE)) %>%
   arrange(avg_delay, p_canc)
```


```{r}
# Ordered overview of average arrival delays per carrier
hflights %>% filter(!is.na(ArrDelay), ArrDelay > 0) %>% group_by(UniqueCarrier) %>% 
  summarise(avg = mean(ArrDelay)) %>% 
  mutate(rank = rank(avg)) %>% 
  arrange(rank)
```

Practice:
```{r}
# How many airplanes only flew to one destination?
hflights %>%
  group_by(Dest) %>%
  summarize(nplanes = n()) %>%
  filter(nplanes == 1) %>%
  summarize(nplanes)

```


```{r}
# Find the most visited destination for each carrier
hflights %>% 
  group_by(UniqueCarrier, Dest) %>% 
  summarise(n = n()) %>% 
  mutate(rank = rank(desc(n))) %>% 
  filter(rank == 1)

```

#### dplyr deals with data with different types

you can use the same functions to data table as in data frame: 

* data table



