---
title: "SleepMammal_example"
output: html_document
---

Sleep Mammal dataset: 

* 11 variables
* 83 observations in total

The columns (in order) correspond to the following:

|column name  |	Description  |
--------------|--------------|
name |	common name
genus |	taxonomic rank
vore |	carnivore, omnivore or herbivore?
order |	taxonomic rank
conservation |	the conservation status of the mammal
sleep_total |	total amount of sleep, in hours
sleep_rem |	rem sleep, in hours
sleep_cycle |	length of sleep cycle, in hours
awake |	amount of time spent awake, in hours
brainwt |	brain weight in kilograms
bodywt |	body weight in kilograms

```{r warning=FALSE, message=FALSE}
library(downloader)
library(tidyverse)
url <- "https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/msleep_ggplot2.csv"
filename <- "msleep_ggplot2.csv"
if (!file.exists(filename)) download(url,filename)
msleep <- read.csv("msleep_ggplot2.csv")
glimpse(msleep)
msleep <- tbl_df(msleep) # change the format for R to display dataframe
```

functions to use:  

* `download(url, filename)`: download the file and give it a file name
* `file.exists(filename)`: returns a logical vector indicating whether the files named by its argument exist. or can use `!file.exists(filename)` for testing if a file does not exist.

# select columns/rows using `select()` or `filter()`

## select columns using `select()`

select name and sleep_total column:  
`select(dataframe, colname1, colname2...)`

```{r}
sleepdata <- select(msleep, name, sleep_total) 
head(sleepdata)
```

use substraction to select cols except some columns:  
`select(df, -colname)`

```{r}
head(select(msleep, -name))
```


select a range of columns by name":"  
`select(df, name1:name2)`

```{r}
head(select(msleep, name:order))
```

select columns names all start with "sl":  
`select(df, starts_with("character"))`

```{r}
head(select(msleep, starts_with("sl"))) # a faster way to gather sleep data
```

Some additional options to select columns based on a specific criteria include

`ends_with()` = Select columns that end with a character string  
`contains()` = Select columns that contain a character string  
`matches()` = Select columns that match a regular expression  
`one_of()` = Select columns names that are from a group of names(in a vector)    
`everything()` = Select every columns

# select rows using `filter()`
filter rows of mammals that sleep a total of more than 16 hours.

```{r}
filter(msleep, sleep_total >= 16)
```

Filter the rows for mammals that sleep a total of more than 16 hours and have a body weight of greater than 1 kilogram.  

```{r}
filter(msleep, sleep_total >= 16, bodywt >= 1)
```

Filter the rows for mammals in the Perissodactyla and Primates taxonomic order:  
`filter(df, colname %in% c("name1", "name2"))`: select rows where only contains name1 and name2 of colname; the same as using "or" logic.

```{r}
a <- filter(msleep, order %in% c("Perissodactyla", "Primates")) 
b <- filter(msleep, order == "Perissodactyla"|order == "Primates") # do the same thing as above
setequal(a,b)
```


# Pipe operator 
This operator allows you to pipe the **output** from one function to the **input** of another function

previously, we select name and sleep_total columns
```{r}
head(select(msleep, name, sleep_total))
```

we can do this by pipe operator as well

```{r}
msleep %>% select(name, sleep_total) %>% head()
```


# arrange or re-order rows using arrange()

arrange the entire dataframe by the order of "order"(colname)  
categorical variable: arranged by alphabetical order A-Z  
numerical variable: arranged from low to high  
we can use `desc()` to reverse the order from high to low or Z-A

```{r}
msleep %>% arrange(order) %>% head # in alphabetical order A-Z
```

we will select three columns from msleep, arrange the rows by the taxonomic order and then arrange the rows by sleep_total.

```{r}
msleep %>% select(name, order, sleep_total) %>% 
  arrange(order, sleep_total) %>%  head() 
# numerical values will be ordered from low to high
```

```{r}
msleep %>% select(name, order, sleep_total) %>% 
  arrange(order, desc(sleep_total)) %>%  
  filter(sleep_total >= 16)
```

# create new columns using `mutate()`
The mutate() function will add new columns to the data frame.  
Create a new column called **rem_proportion** which is the ratio of rem sleep to total amount of sleep.

```{r}
msleep <- msleep %>% mutate(rem_proportion = sleep_rem / sleep_total)
head(msleep)
```

Here we add a second column called **bodywt_grams** which is the bodywt column in grams.

```{r}
msleep <- msleep %>% mutate(bodywt_grams = bodywt * 1000)
head(msleep)
```

# create summaries of data frame using `summarise()`

The `summarise()` function will create summary statistics(a new table) for **a given column** in the data frame such as finding the mean. For example, to compute the average number of hours of sleep, apply the mean() function to the column sleep_total and call the summary value avg_sleep.

```{r}
msleep %>% summarise(avg_sleep = mean(sleep_total))
```

other commonly used functions

```{r}
a <- msleep %>% summarise(avg_sleep = mean(sleep_total),
                     min_sleep = min(sleep_total),
                     max_sleep = max(sleep_total),
                     sd_sleep = sd(sleep_total),
                     med_sleep = median(sleep_total),
                     sum_sleep = sum(sleep_total),
                     first_sleep = first(sleep_total),
                     last_sleep = last(sleep_total),
                     distinct_order = n_distinct(order),
                     total = n())
```

# Group operation + summarise
## Split - apply - combine 

* split the msleep data frame by the taxonomic order;  
* summarize the max, min, and avg of total sleep time for each order;  
* combine these data together
```{r}
msleep %>% group_by(order) %>% 
  summarise(avg_sleep = mean(sleep_total),
            max_sleep = max(sleep_total),
            min_sleep = min(sleep_total),
            total = n()) # how many samples from each order
```

