---
title: "ggvis cookbook"
output: html_document
---

```{r message=FALSE, warning=FALSE}
library(ggvis)
library(dplyr)
mtcars
```

### scatterplots
mtcars: plot  `wt`(x) against `mpg`(y) using scatterplot and add a smooth trend line; color indicate `cyl`.
```{r}
mtcars %>% ggvis(~wt, ~mpg, fill = ~cyl) %>% layer_points() %>% layer_smooths()
```

use more options in `layer_points`: 

* `size` of dot, `shape` 
* `stroke` as the color of a point
* `fill := NA` means do not fill the dot point with color.

```{r}
mtcars %>% 
  ggvis(~wt, ~mpg) %>%
  layer_points(size := 30, shape := "diamond", stroke := "blue", fill := NA)
```

more options in add trend line  
plot a regression line with 95% confidence in the model
```{r}
mtcars %>% 
  ggvis(~wt, ~mpg) %>%
  layer_points() %>%
  layer_model_predictions(model = "lm", se = TRUE) # se allows prediction interval
```

### scatter plot with grouping
coloring point by `cyl`  

* first convert `cyl` to categorical variable using factor
* fill the scatter plot by this factor

```{r}
mtcars %>%
  ggvis(~wt, ~mpg) %>%
  layer_points(fill = ~factor(cyl), size := 50)
```


* color points by `cyl` factors
* add smooth trend line to each group of `cyl`: can only achieved after `group_by` by cyl
```{r}
mtcars %>%
  ggvis(~wt, ~mpg, fill = ~factor(cyl)) %>%
  layer_points() %>%
  group_by(cyl) %>%
  layer_model_predictions(model = "lm")
```

### bar graphs
use pressure data frame
```{r}
head(pressure)
```

plot `temprature`(x) vs `pressure`(y) using bar graphs, both are continuous var
```{r}
pressure %>%
  ggvis(~temperature, ~pressure) %>%
  layer_bars(width = 10) # speicify the width of each bar
```

transform `temprature`(x) into factors, categorical variables and plot again
```{r}
# change the col of temp into a col of factor temprature
pressure2 <- pressure %>% mutate(temperature = factor(temperature))
head(pressure2)

pressure2 %>% 
  ggvis(~temperature, ~pressure) %>%
  layer_bars()

```

### line graphs
lines with points
```{r}
pressure %>% ggvis(~temperature, ~pressure) %>% layer_lines() %>% layer_points()
```

### histograms
```{r}
head(faithful)
```

basic histograms
```{r}
faithful %>% ggvis(~eruptions) %>% layer_histograms()
```

bin selection
```{r}
faithful %>% ggvis(~eruptions) %>% layer_histograms(width=0.5, fill := "#fff8dc", 
                                                    boundary = 0)

```


By default, when the number of integer values is small, bins will be centered at integers and have a width of 1: 
```{r}
head(cocaine)
cocaine %>% ggvis(~month, fill := "#fff8dc") %>%
  layer_histograms() %>%
  add_axis("x",title = "month") %>%
  add_axis("y", title = "count")
```


### box plots
```{r}
mtcars %>% ggvis(~factor(cyl), ~mpg) %>% layer_boxplots()
```

