---
title: "tidy missing data"
author: "Eleanor Zhang"
output:
  html_document: 
    theme: lumen
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE)
```

Load packaged at need.
```{r}
library(tidyverse)
library(naniar) # deal with missing value
```

# Search and replace missing values

Missing data are not always coded as NA! The objective here is to:  

*   Look for hidden missing values
*   Replace missing values with NA
*   Check assumptions on missingness

Some possible incorrectly coded missing values: "missing", "Not Available", "N/A"... These need to be recoded into "NA".

Let's create a chaotic case
```{r}
bad.df <- tibble(score = c(3, -99, 4, -99, 7, 10, 12, 16, 9),
            grade = c("N/A", "E", "missing", "na", "n/a", "NULL", ".", "NULL", "N/a"),
            place = c(-99, 97, 95, 92, -98, "missing", 88, ".", 86))
```

## search for missing values

Seach for specific entry and it returns number of occurence in each variable
```{r}
bad.df %>% miss_scan_count(search = list("N/A"))

# it can take multiple entries to search
bad.df %>% miss_scan_count(search = list("N/A", "N/a"))
```

## Replace targeted entries with NA
```{r}
# replace specific entries in grade with NA
bad.df %>% replace_with_na(replace = list(grade = c("N/A", "N/a")))
```

To reduce code repetition and provide other options to replace with NA using scope of variants:  

*    `replace_with_na_all()`: all variables
*    `replace_with_na_at()`: a subset of variables
*    `replace_with_na_if()`: replace a subset of variables that satisfied some condition

Replace all entries of -99 with NA
```{r}
bad.df %>% replace_with_na_all(condition = ~.x == -99)
```

Replace some entries with NA
```{r}
bad.df %>% replace_with_na_all(condition = ~.x %in% c("N/A", "missing", "na"))
```

Replace in a subset of variables
```{r}
replace_with_na_at(bad.df, 
                   .vars = c("score", "grade"),
                   ~.x %in% c("N/A", "missing", "na"))
```

Replace NA in variables that are characters
```{r}
replace_with_na_if(bad.df, 
                   .predicate = is.character,
                   ~.x %in% c("N/A", "missing", "na"))
```

# Implicit missing values

These values are implictly missing without explicitly listed.

First contruct a case: values recorded in the morning, afternoon and evening. By construction, Sam miss score in evening (it's not there...)
```{r}
set.seed(1)
bad.df2 <- tibble(name = c(rep("robin", 3), rep("sam", 3), rep("blair",3)),
              time = rep(c("morning", "afternoon", "evening"), 3),
              value = sample(100:1000, 9)) %>% slice(-6)
bad.df2
```

## reveal implicit missings

To reveal this missingness, we can spread the dataframe from long format to wide format
```{r}
spread(bad.df2, key = time, value = value)
```

To summarize, there are two types of missing values:  

*   Explicitly: missing with NA
*   Implicitly: Not shown in the data, but implied

## complete and fill missings

Use `complete()` from tidyr package to make implicit missings explicit:
```{r}
df2.complete <- bad.df2 %>% tidyr::complete(name, time) # a combination of variables
df2.complete
```

More often than not, a readable tabular is not easy to deal with in silico:
```{r}
df.readable <- df2.complete %>% mutate(name = ifelse(duplicated(name), NA, name))
df.readable
```

In this case, we can fill down the column by `fill()` from tidyr. This situation is called "lost information carried forward" (LOCF).
```{r}
df.readable %>% fill(name)
```

We can execute fill() and complete() in one pipeline
```{r}
df.readable2 <- bad.df2 %>% mutate(name = ifelse(duplicated(name), NA, name))
df.readable2

df.readable2 %>% fill(name) %>% complete(name, time)
```


__Remark__: These techniques need to be handled very carefully because it only solves problems in some restricted situation.

# Missing data dependence

When facing missing data, we have to carefully decide either impute the missings or simply delete them.There are three types of missingness generally and first two are easily confusing:

*   MCAR: missing completely at random
*   MAR: missing at random
*   MNAR: missing not at random

## missing completely at random (MCAR)

The missings are unrelated to any observed or unobserved data. It means the missings is a random subset of all observations. Statistically, it means the imaginery distribution of missings in a given variable is similar to those observed ones. 

If this is true, then we should:  

*    Impute the data
*    Delete with careful attention on reduced sample size, but no bias will be introduced with deletion. Practically it is adviced do not delete the data unless they constitute less than 5% of the data.

## missing at random (MAR)

MAR is a weaker assumption than MCAR. This is a conditional randomness, meaning the missings can be related to observed data but are not allowed to relate to unobserved data. The missings can be explained by some observed covariates. For example, suppose we have measured blood pressure for patients. Within males or females (conditional on sex), the distribution of missing blood pressure are similar to those observed ones. 

If this is true, we should:  

*   Impute data carefully
*   Better not delete missings because deletion will introduce bias due to its dependence on observed covariate.

More explicit discussion about these two types of missingness can be found in this short [paper](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4121561/)

## missing not at random (MNAR)

Missingness is related to the variable itself. For example, people with bad medical condition are more likely to drop out of the study than those healthy ones.

__Remark__: Whether such missingness is random or not is not testable based observed data. But it is can be infered by prior knowledge. For example, we may suspect neuropsyhological test score may be more likely to miss among people with intellectual disability because they cannot complete the test due to pre-existing condition. Certainly this is not MCAR, but it is still possible to be MAR if within given intellectual level, the missings are random. Therefore, the concept

# Visualize missing pattern

Try out some rearrangement of the dataset and visualize the missingness can be a good starting point of looking for missing randomly or not, although this is not the definitive technique to use. Here we use the `msleep` dataset from ggplot2 package.

```{r}
head(msleep)
```

Visualize missingness
```{r}
vis_miss(msleep) 
```

Comment: It looks like random missing. We can try rearrange by covariate "genus". 
```{r}
msleep %>% arrange(genus) %>% vis_miss()
```

Comment: Still no obvious pattern.

Visualize any MNAR by turning on "cluster".
```{r}
vis_miss(msleep, cluster = TRUE)
```

Summary:  

*    If the missing is MCAR: None of the variable observed/unobserved will explain the missingness.
*    If missing is MAR: some observed variable can explain missingness, meaning you will see cluster of missingness if you arrange the entire dataset by those variables.
*    If missing is MNAR: you will see a bisection of dataset by certain variables.